### 2009-06-16 paper.tex
last edits

@@

\begin{description}
\item{\stemloc:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
{\bf This model improves on the previous model (Long Indel) by introducing RNA secondary structure, though it no longer models phylogeny as rigorously as the Long Indel model} (it uses single-linkage clustering instead of a true phylogeny).
\item{\stemlocama:} the \stemloc\ program was used to align the three sequences in ``sequence annealing'' mode,
 a posterior decoding method that attempts to optimize AMA, a sum-over-pairs alignment accuracy metric \cite{BradleyPachterHolmes2008}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
{\bf This program uses the previous model but instead of maximizing likelihood, it attempts to maximize an alignment accuracy metric.}
\item{TKF91:} with the TKF91 model \cite{ThorneEtal91}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied (as they were for the \indiegram\ benchmark).
The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
{\bf This may be understood as a naive reconstruction that completely ignores the stem sub-model of TKFST.}
\item{Long Indel:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied.
The deletion and substitution rates were set equal to those of the loop submodel of TKFST.
The mean indel length was set equal to $D_l$, the mean number of bases that are created/removed by an insertion/deletion in the loop submodel of TKFST;
the mean equilibrium sequence length (and thereby the insertion rate) was equal to $B_l$, the mean number of bases in TKFST at equilibrium
(\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
{\bf This model improves on the previous model (TKF91) by introducing affine gap penalties.}
\end{description}

We focused primarily on alignment accuracy, under the simplifying assumption that this will be a good indicator for
(and perhaps the primary determinant of) the accuracy of ancestral reconstruction.

We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
Theory predicts that ML inference, using the correct model and parameters, should maximize the perfect alignment rate.
Inspecting \figref{perfect}, we find this to be the case, except where the outgroup is very distant and the bins are probably undersampled
(when the perfect alignment rate drops below $\sim 5/125$, 125 trials may be too few).
We also note that the ML version of \stemloc\ is near-optimal, despite the \stemloc\ pair-SCFG being
slightly different from the TKFST pair-SCFG in parameterization and structure
 (e.g. \stemloc's grammar does not allow insertion/deletion of entire substructures).
The ML version of \stemloc\ is also observed to have a slightly higher perfect alignment rate than the posterior-decoding version (\stemloc-AMA).
Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods,
with no decisive advantage shown by either TKF91 (linear gap penalty) or Long Indel (affine gap penalty).

@@

Taken together, these results show that the most important factor distinguishing the various models
is the incorporation of some form of basepair structure:
structure-less \handel\  (regardless of linear {\em vs} affine gap penalty) performs much worse than the structure-aware SCFG methods.
Intuitively, this is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
correctly aligned for free.

The second most important, amongst the factors we have considered in this experiment, is use of the correct posterior-decoding algorithm for the objective function at hand
(c.f. \stemloc\ {\em vs} \stemlocama).
An obvious corollorary is the importance of choosing an appropriate objective function in the first place
(c.f. perfect alignment rate {\em vs} AMA).

Lastly, the fine details of the scoring scheme---including branch lengths, substitution scores, gap penalties and so forth---appear to
be the least important of the factors we considered, yielding observable differences only when all other aspects of the inference procedure were more-or-less equal.

@@

\begin{figure}[!ht]
  \centering
%  \includegraphics [scale=0.4] {figs/parent.pdf}
   \caption{
     \textbf{A simple example of transducer composition to build an
       evolutionary model of two extant sequences.}
    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
     and two branch transducers (the gray boxes labeled $T_X$ and $T_Y$)
     mutate it according to the specified evolutionary model.
     Gray nodes correspond to observed data and white nodes unobserved data.
   }
   \figlabel{parent}
 \end{figure}

 \begin{figure}[!ht]
   \centering
 %  \includegraphics [scale=0.4] {figs/threeway.pdf}
   \caption{
     \textbf{Transducer composition on a star phylogeny with three
       (extant) leaf sequences.}
     An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
     and three branch transducers (the gray boxes labeled $T_X$, $T_Y$ and $T_Z$)
     mutate it according to the specified evolutionary model.
     Gray nodes correspond to observed data and white nodes unobserved data.
     If the branch transducers are time-reversible, then this star phylogeny with three leaves is
     the neighborhood of any interior node in a (binary) phylogenetic tree, 
     from which it follows that evaluating the likelihood function
     on this star phylogeny is sufficient, in principle, for a sampling algorithm on an arbitrary phylogeny.
   }
\figlabel{threeway}
 \end{figure}

@@

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/perfect.pdf}
  \caption{
    \textbf{Dependence of perfect alignment rate on outgroup branch
      length.}
    We investigated the perfect alignment rate
    (the number of times that a given alignment program estimates the alignment 100\% correctly, with no errors)
    as a function of outgroup branch length, simulating the
    evolution of three structural RNAs under our model.  The two
    sister species were equidistant from the ancestral sequence, with
    a branch length of 1, corresponding to one expected
    substitution per site in loop sequence.  We allowed the outgroup
    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
    Since the perfect alignment rate is rather low, we further aggregated the $t$-values for this plot into bins of five;
    thus, for example, the bin named ``$0 to 0.4$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$.
    The vertical axis counts the number of perfect alignments in each bin, out of a total of 125 trials per bin.
}
  \figlabel{perfect}
\end{figure}

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/ama.pdf}
  \caption{
    \textbf{Dependence of reconstruction accuracy on outgroup branch
      length.}
    We investigated the dependence of structural reconstruction and
    alignment accuracy on outgroup branch length by simulating the
    evolution of three structural RNAs under our model.  The two
    sister species were equidistant from the ancestral sequence, with
    a branch length of 1, corresponding to one expected
    substitution per site in loop sequence.  We allowed the outgroup
    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
    The Alignment Metric Accuracy (AMA) is, roughly, the proportion of residues that are correctly
    aligned, averaged over all pairs of sequences
    (see \cite{SchwartzMyersPachter2006} for a precise definition; we used AMA with Gap Factor 1).
}
  \figlabel{ama}
\end{figure}

>>>>

The above-described TKFST SCFGs must be transformed slightly before they can be loaded into \indiegram.
The grammars are presented in \indiegram\ format in \supptext{4}.

@@

\begin{description}
\item{\stemloc:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
Instead of modeling a true phylogeny by introducing unobserved ancestral sequences, it just does single-linkage clustering of the observed sequences.
\item{\stemlocama:} the \stemloc\ program was used to align the three sequences in ``sequence annealing'' mode,
 a posterior decoding method that attempts to optimize AMA, a sum-over-pairs alignment accuracy metric \cite{BradleyPachterHolmes2008}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
This program uses the same underlying pair SCFG as \stemloc, but instead of maximizing likelihood, it attempts to maximize an alignment accuracy metric.
\item{TKF91:} with the TKF91 model \cite{ThorneEtal91}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied (as they were for the \indiegram\ benchmark).
The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
This may be understood as a naive sequence-only reconstruction that completely ignores basepair structure (i.e. the stem sub-model of TKFST).
\item{Long Indel:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied.
The deletion and substitution rates were set equal to those of the loop submodel of TKFST.
The mean indel length was set equal to $D_l$, the mean number of bases that are created/removed by an insertion/deletion in the loop submodel of TKFST;
the mean equilibrium sequence length (and thereby the insertion rate) was equal to $B_l$, the mean number of bases in TKFST at equilibrium
(\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
This model improves on the previous model (TKF91) by introducing affine gap penalties.
\end{description}

We measured alignment accuracy, under the simplifying assumption that this correlates well with ancestral reconstruction accuracy.

We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
Theory predicts that Maximum Likelihood inference, using the correct model and parameters, should be the algorithm that scores the most perfect estimates.
Inspecting \figref{perfect}, we find this to be the case, except where the outgroup is very distant and the bins are probably undersampled
(i.e. the departure from prediction that we observe for low-identity alignments is not statistically significant:
when the perfect alignment rate drops below $\sim 5/125$, then 125 trials are probably too few).
We also note that the ML version of \stemloc\ is near-optimal, despite the \stemloc\ pair-SCFG being
slightly different from the TKFST pair-SCFG in parameterization and structure
 (e.g. \stemloc's grammar does not allow insertion/deletion of entire substructures).
The ML version of \stemloc\ is also observed to have a slightly higher perfect alignment rate than the posterior-decoding version (\stemloc-AMA).
Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods;
furthermore, both linear (TKF91) and affine (Long Indel) gap-penalties perform equally bad in this test
(note that the TKFST model, from which the true alignments were simulated, does not allow long-indel events,
which may partly explain why affine gap-penalties do not help in this benchmark).

@@

Taken together, these results suggest that the most important factor distinguishing the various models we have examined
is the incorporation of some form of basepair structure:
structure-blind \handel\  (regardless of linear {\em vs} affine gap penalty) performs much worse than the structure-aware SCFG methods.
Intuitively, this is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
correctly aligned for free.
In benchmarks of RNA multiple alignment programs, structure-aware scoring schemes routinely outperform structure-blind scoring schemes \cite{GardnerEtAl2005,WilmEtAl2006}.
Since we know that modeling structure is very important, it's not too surprising that it turns out to be the most important of the factors we considered.

The second most important, amongst the factors we have considered in this experiment, is
selection of the most appropriate objective function for the task at hand
(c.f. perfect alignment rate {\em vs} AMA),
followed by use of the correct posterior-decoding algorithm for the chosen objective function
(c.f. \stemloc\ {\em vs} \stemlocama).
This is a subtle but important point: before deciding exactly what inference algorithm we're going to use to reconstruct ancestral sequences,
we need to decide whether we want to maximize
(a) the probability that our reconstructed sequence is 100\% correct,
(b) the expected number of nucleotides that are correctly reconstructed,
(c) the expected number of base-pairs that are correctly reconstructed,
(d) the expected number of stems that are correctly reconstructed,
(e) some other metric.
Each of these metrics would require a slightly different inference algorithm.

Lastly, the fine details of the scoring scheme---including branch lengths, substitution scores, gap penalties and so forth---appear to
be the least important of the factors we considered, yielding observable differences only when all other aspects of the inference procedure were more-or-less equal.
While such details of the model may affect reconstruction quality,
they appear to have very minor influence on alignment quality.

@@

@@

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/perfect.pdf}
  \caption{
    \textbf{Dependence of perfect alignment rate on alignment method and outgroup branch
      length.}
    Perfect alignment is when a given alignment program estimates the alignment 100\% correctly, with no errors.
    Simulating the evolution of three structural RNAs under the TKFST model,
    we investigated the dependency of perfect alignment rate on outgroup branch length.
    The simulation included two sister species at unit distance from the ancestral sequence, plus one ``outgroup''
    whose branch length $t$ was varied between $[0, 2.5]$ by selecting 25 equally spaced values of $t$ in this range, spaced $0.1$ apart.
    (A unit-length branch here corresponds to one expected substitution per site in loop sequence.)
    We simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
    Since the perfect alignment rate is rather low, we further aggregated the $t$-values into bins of five;
    thus, for example, the bin named ``$0 to 0.4$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$ and represents $25 \times 5 = 125$ trials in total.
    The perfect alignment rate was measured for various statistical alignment inference procedures.
    These procedures are described in the text, but may be summarized very briefly as
    ML under the true model (``\indiegram'');
    greedy approximate-ML progressive alignment by single-linkage clustering with pair SCFGs (``\stemloc'');
    sequence annealing, a form of posterior decoding to maximize a sum-over-pairs accuracy metric, using pair SCFGs to get the posterior probabilities (``\stemlocama'');
    statistical alignment using the TKF91 model, i.e. linear gap-penalties (``TKF91''); and
    statistical alignment using a long-indel model, i.e. affine gap-penalties (``Long Indel'').
}
  \figlabel{perfect}
\end{figure}

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/ama.pdf}
  \caption{
    \textbf{Dependence of alignment metric accuracy on alignment method and outgroup branch
      length.}
    We simulated the evolution of three structural RNAs under the TKFST model.
    The simulation included two sister species at unit distance from the ancestral sequence, plus one ``outgroup''
    whose branch length $t$ was varied between $[0, 2.5]$ by selecting 25 equally spaced values of $t$ in this range, spaced $0.1$ apart.
    We then simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
    The Alignment Metric Accuracy (AMA) is, roughly, the proportion of residues that are correctly
    aligned, averaged over all pairs of sequences
    (see \cite{SchwartzMyersPachter2006} for a precise definition; we set the AMA Gap Factor to 1).
    The AMA between the true alignment and the inferred alignment was measured for various statistical alignment inference procedures.
    These procedures are described in the text, but may be summarized very briefly as
    ML under the true model (``\indiegram'');
    greedy approximate-ML progressive alignment by single-linkage clustering with pair SCFGs (``\stemloc'');
    sequence annealing, a form of posterior decoding to maximize a sum-over-pairs accuracy metric, using pair SCFGs to get the posterior probabilities (``\stemlocama'');
    statistical alignment using the TKF91 model, i.e. linear gap-penalties (``TKF91''); and
    statistical alignment using a long-indel model, i.e. affine gap-penalties (``Long Indel'').
}
  \figlabel{ama}
\end{figure}

### 2009-06-16 s3.tex
last edits

\subsection{Automated grammar construction}
We implemented our model construction algorithm on the three-taxon star phylogeny
in a set of Perl scripts.
Given a singlet transducer modeling ancestral sequences and structures and
a branch transducer modeling structural evolution,
the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
All possible models of structural evolution which can be represented by a Pair SCFG
are permitted as input to the scripts,
allowing for flexible and automated model design.

>>>>

 \begin{figure}[!ht]
   \centering
  \includegraphics [scale=0.4] {figs/threeway.pdf}
   \caption{
     \textbf{Transducer composition on a star phylogeny with three
       (extant) leaf sequences.}
     An ancestral sequence $W$ evolves into three descendant sequences $X$, $Y$ and $Z$.
     A singlet transducer (the horizontal gray box) emits an ancestral sequence and structure
     and three branch transducers (the gray boxes labeled $T_X$, $T_Y$ and $T_Z$)
     mutate it according to the specified evolutionary model.
     Gray nodes correspond to observed data and white nodes unobserved data.
     If the branch transducers are time-reversible, then this star phylogeny with three leaves is
     the neighborhood of any interior node in a (binary) phylogenetic tree, 
     from which it follows that evaluating the likelihood function
     on this star phylogeny is sufficient, in principle, for a sampling algorithm on an arbitrary phylogeny.
   }
\figlabel{threeway}
 \end{figure}

\subsection{Automated grammar construction}
We implemented our model construction algorithm on the three-taxon star phylogeny (\figref{threeway}) 
in a set of Perl scripts.
Given a singlet transducer modeling ancestral sequences and structures and
a branch transducer modeling structural evolution,
the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
All possible models of structural evolution which can be represented by a Pair SCFG
are permitted as input to the scripts,
allowing for flexible and automated model design.

### 2009-06-16 s4.tex
last edits

\newpage
\section{Software}

All tools are available from \darturl\ as part of the \dart\ software package for sequence analysis.

\subsection{Automated grammar construction}
We implemented our model construction algorithm on the three-taxon star phylogeny
in a set of Perl scripts.
Given a singlet transducer modeling ancestral sequences and structures and
a branch transducer modeling structural evolution,
the scripts generate C++ code to create the corresponding jointly-normalized SCFG.
All possible models of structural evolution which can be represented by a Pair SCFG
are permitted as input to the scripts,
allowing for flexible and automated model design.

Given files describing the singlet and branch transducers,
including weights of all transitions which may be functions of evolutionary distance,
the package ComposedTreeTransducer::FourWayComposedTT can automatically generate
the state graph and transition matrix of a multi-sequence model of three extant sequences.
It removes the useless windback $\Snull$ states described in the paper and
introduces effective direct transitions caused by bifurcations with possibly-empty children.
The package ComposedTreeTransducer::TripletSCFG transforms a multi-sequence
model created by ComposedTreeTransducer::FourWayComposedTT into the 
corresponding jointly-normalized three-sequence SCFG
and generates C++ code to build the model.

Example singlet and branch transducers files are provided for a simple 
Pair HMM model, a simple Pair SCFG model and the full TKF Structure Tree model.

\subsection{Reconstruction of ancestral structures}
The program \indiegram\ can perform maximum-likelihood inference on the
three-sequence SCFGs automatically generated by the ComposedTreeTransducer::TripletSCFG package.
Complete or no structural information for the three extant sequences can be supplied
as input.

\subsection{Simulation of RNA family evolution}
The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, in combination with the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.

>>>>

\newpage
\section{Inference grammars for the TKFST model}
\seclabel{TKFST}

This section describes multi-sequence grammars that can be used to infer alignments, structures and ancestors under the
TKF Structure Tree (TKFST) model.

\input{tkfst.tex}

\newpage
\section{The TKF Structure Tree model as a transducer} \seclabel{tkfst}

We can cast the TKF Structure Tree (TKFST) model as a transducer.
This means we can automatically deduce rules like those shown for the TKFST model in \secref{TKFST}.

\subsection{The single-sequence TKFST model as a singlet transducer}

The states of the singlet transducer are shown in \tabref{tkfstsinglet}.
Allowed transitions between these states are shown in \secref{TKFST}.

\begin{table}[!ht]
  \centering
  \begin{tabular}{lrrrr}
    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
    $L$ & $\Sstart$ \\
    $I_L$ & $\Sinsert$ & & $(x,\Tnull)$ & $p(x)$ \\
    \\
    $S$ & $\Sstart$ \\
    $I_S$ & $\Sinsert$ & & $(x,y)$ & $p(x,y)$ \\
    \\
    $B$ & $\Sinsert$ & & $(L\,S)$ & 1 \\
  \end{tabular}
  \caption{
    \tablabel{tkfstsinglet}
    States of the singlet transducer of the TKF Structure Tree model \cite{Holmes2004}.
    Singlet transducers can only have states of type $\Sstart$ or $\Sinsert$.
    This is the indiegram-style transducer equivalent of the singlet TKFST SCFG in \secref{TKFST}.
  }
\end{table}

\subsection{The two-sequence TKFST model as a branch transducer}

The states of the branch transducer are shown in \tabref{tkfstbranch}.
Allowed transitions between these states are shown in \secref{TKFST}.

\begin{table}[!ht]
  \centering
  \begin{tabular}{lrrrr}
    state & $\type$ & $\absorb$ & $\emit$ & $e(\,\bullet\,|\mathrm{TKFST})$ \\ \hline
    $L$ & $\Sstart$ \\
    $I_L$ & $\Sinsert$ & & $(u,\Tnull)$ & $p(u)$ \\
    $M_L$ & $\Smatch$ & $(x,\Tnull)$ & $(u,\Tnull)$ & $p(u|x)$ \\
    $D_L$ & $\Smatch$ & $(x,\Tnull)$ & $(\Tgap,\Tnull)$ & 1 \\
    $W_L$ & $\Swait$ \\
    \\
    $S$ & $\Sstart$ \\
    $I_S$ & $\Sinsert$ & & $(u,v)$ & $p(u,v)$ \\
    $M_S$ & $\Smatch$ & $(x,y)$ & $(u,v)$ & $p(u,v|x,y)$ \\
    $D_S$ & $\Smatch$ & $(x,y)$ & $(\Tgap,\Tgap)$ & 1 \\
    $W_S$ & $\Swait$ \\
    \\
    $B_i$ & $\Sinsert$ & & $(L_i\,S_i)$ & 1 \\
    $B$ & $\Smatch$ & $(L\,S)$ & $(L\,S)$ & 1 \\
    $B_p$ & $\Smatch$ & $(L\,S)$ & $(L\,\Send)$ & 1 \\
    $B_e$ & $\Smatch$ & $(L\,\Send)$ & $(L\,\Send)$ & 1 \\
  \end{tabular}
  \caption{\tablabel{tkfstbranch}
    States of the branch transducer of the TKF Structure Tree model \cite{Holmes2004}.
    States which have the same names as states of the singlet transducer in \tabref{tkfstsinglet}
    are the branch equivalent of the corresponding singlet states
    (e.g. a $\Smatch$ state might be the branch equivalent of an $\Sinsert$ state).
    States $L_i$ and $S_i$ are the $\Sstart$ states of a sub-model (not shown) identical
    in structure to the singlet transducer.  They are used to insert a new stem-loop structure.
    This is the indiegram-style transducer equivalent of the conditional pair TKFST SCFG in \secref{TKFST}.
  }
\end{table}

\subsection{The multi-sequence TKFST model as a composite transducer}

We can use the state graph construction algorithm described 
in the paper and detailed in Text S1 to create a model of the simultaneous evolution
of several sequences.

\begin{figure}[!htb]
  \centering
  \includegraphics [scale=0.4] {figs/parent.pdf}
  \caption{\figlabel{parent}
    A simple example of transducer composition to build a multi-sequence model of two extant sequences.
    An ancestral sequence $W$ evolves into two descendant sequences $X$ and $Y$.
    A singlet transducer (the horizontal gray box) emit ancestral sequence and structure
    and two branch transducers (the gray boxes labeled $\Delta T_X$ and $\Delta T_Y$)
    mutate it according to the specified multi-sequence model.
    Gray nodes correspond to observed data and white nodes unobserved data.
  }
\end{figure}

Consider the simple model shown in \figref{parent}.
The state of the multi-sequence model describing this model is a 3-vector $\bvec{a} = \left( a_1,\,a_2,\,a_3 \right)$,
where $a_1$ is the state of the singlet transducer generating the ancestral sequence $W$
and $a_2$ and $a_3$ are the states of the branch transducers evolving $W$ into
extant sequences $X$ and $Y$.

We can show some of the allowed transitions of this multi-sequence model.
The state of the branch transducer associated with the active node $n$ is shown in bold.

\paragraph{Stem creation.}

A stem is created at the root node $W$ (corresponding to a bifurcation in the singlet transducer $a_1$)
and survives in the sequence $X$ at node 2 but is deleted in the sequence $Y$ at node 3.

\begin{align}
  \nakedthreevec{1:}{2:}{3:} \quad
  \threevec{L}{L}{\mathbf{L}} &
  \to \threevec{\mathbf{L}}{W_L}{W_L}
  \to \threevec{B}{B}{B_p}
  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{S}{\mathbf{S}}{\Send}
  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\mathbf{S}}{W_S}{\Send} \\
  & \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\mathbf{S}}{W_S}{\Send}
\end{align}

\paragraph{Stem insertion.}

A stem sequence is inserted in sequence $X$ at node 2.

\begin{align}
  \nakedthreevec{1:}{2:}{3:} \quad
  \threevec{L}{L}{\mathbf{L}} \to \threevec{L}{\mathbf{L}}{W_L}
  \to \threevec{L}{B_i}{B}
  \to \threevec{L}{L}{\mathbf{L}} \, \threevec{\Send}{S}{\mathbf{S}}
  \to \threevec{\mathbf{L}}{W_L}{W_L} \, \threevec{\Send}{\mathbf{S}}{W_S}
\end{align}

\paragraph{Stem termination.}

All stem sequences are ended by (possibly empty) loop sequences.

\begin{align}
  \nakedthreevec{1:}{2:}{3:} \quad
  \threevec{\mathbf{S}}{W_S}{W_S}
  \to \threevec{B_e}{B_e}{B_e}
  \to \threevec{\Send}{\Send}{\Send} \, \threevec{L}{L}{\mathbf{L}}
  \to \threevec{\Send}{\Send}{\Send} \, \threevec{\Send}{\Send}{\Send}
\end{align}

The functions $\alpha(t)$, $\beta(t)$ and $\gamma(t)$ are parametrized by the insertion and deletion rates of the Structure Tree model.
They are defined for loop sequences as
\begin{align}
  \kappa_1 &= \lambda_1 / \mu_1 \\
  \alpha_1 &= \exp \left( -\mu_1 t \right) \\
  \beta_1 &= \frac{\lambda_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) } \\
  \gamma_1 &= 1 - \frac{\mu_1 \left( 1 - \exp \left((\lambda_1 - \mu_1) t \right) \right)}{\left( 1 - \exp (- \mu_1 t) \right) \left(\mu_1 - \lambda_1 \exp \left( (\lambda_1 - \mu_1) t \right) \right) }
\end{align}
and similarly for stem sequences \cite{Holmes2004}.
